Numerical trajectory patterns of floating macroalgae

### 2021-03-15 thesis_chapter_3.Rmd
round 1 additions to chapter 3 writing done

Past research has inferred from direct and indirect techniques that drifting macroalgae tend to follow the prevailing surface currents [@thiel2005; @hobday2000; @fraser2011; @rothausler2011; @rothausler2011]. The results from this numerical experiment confirm that drifting macroalgae do follow the surface current in the study region when considering only hydrodynamic drag. In addition, not only are the trajectories similar but so are the end points at the end of simulation. However, when **ad-hoc** wind drag (windage) is considered in combination with hydrodynamic drag the end locations differ greatly when compared to purely Lagrangian particle trajectories. The inclusion of wind drag causes the particles to flow further away from the coast and with a greater distance travelled from the release location. The difference in trajectory and distance covered can be attributed to the inclusion of wind drag. Wind drag adds to the effects of hydrodynamic drag when the wind direction and current direction are opposite to each other. However, when the wind direction is in the same or similar direction to that of the current, momentum energy from the wind is transferred to the object [@hackett2006; @putman2020]. This causes the particles to travel further as well as exposed to different time varying flows. Past studies have shown this when including windage as a function of inertia [@putman2018;@brooks2019; @putman2020]. A study by @putman2020 used ad-hoc windage factors that was based on the surface current velocity to assess the approach for improving transport predictions of pelagic Sargassum. The results showed that including ad-hoc windage factors improved virtual trajectories compared to tracked Sargassum mats which was partly due to the inclusion of momemtum energy transfer from wind. This is also reflected in the analysis comparing distances travelled by all the particles in each simulation. The particles in the simulations that included wind drag traveled significantly greater distances compared to the passive or hydrodynamic only drag.

The addition of drag forces causes the particles to cluster together along the mean trajectory compared to that of the passive particles which are more evenly dispersed. This is most likely due to the shared drag characteristics among the particles which included the plants weight and surface area into the calculation. Other studies investigating flotsam trajectory characteristics have found similar results [@breivik2011; @olascoaga2020; @miron2020]. A study by @miron2020 conducted field experiments using a range of objects such as spheres, cubes, cuboids to compare to investigate the effect of inertia on particles dynamics. The results from the study showed that objects tended to cluster according to the shape of the object which the authors attributed to shared characteristics of inertia.

The ocean is made up of varying time-varying flows and changes in an objects velocity can cause the object to be exposed to different flow patterns over time. The lack of dissimilarity in trajectory between different cross-sectional area types (minimum, mean and maximum) suggests that reduction in velocity in relation to cross-sectional area is negligible. A similar result was found by @gouvello2020 who investigated the effects of swimming behavior on sea turtle hatchling dispersal in the Agulhas region. The results suggested that the hatchlings trajectories are mostly influenced by the ocean currents in the first year of hatching due to  as juveniles do are not swim fast enough to alter thier overall trajectory. Therefore, the differences in cross-sectional area types may not be significant enough in order to have an effect on the relevant velocity vectors (zonal and meridonal). Another possible reason is the resolution of the ocean model used. The resolution of the underlying ocean model is an important aspect of Lagrangian ocean modeling and must be able to resolve sub-grid scale ocean processes for finer scale applications. A study by @hart2018 assessed the inclusion of stochastic motion, wind and currents into forecasting for search and rescue using the same model ocean model used in this study. The authors found that the inclusion of brownian motion greatly increased the accuracy in representing sub-grid scale processes and that the inclusion of wind, currents and stochastic motion greatly improved forecasting applications. Therefore, the authors argue that the lack of dissimilarity between simulations of different cross-sectional area and drag exposure scenarios is not as a result of the resolution of the model.

The same is also true when comparing different hydrodynamic and wind drag exposure scenarios. These scenarios reflect different buoyancy situations which result in different amounts of surface area exposed to hydrodynamic and/or wind drag. The similarities between the different combined drag scenarios suggests that higher magnitudes of wind exposure do not significantly alter trajectories. Instead, the results from this study suggest that the inclusion of drag forces in simulating macroalgal trajectory may result in improving accuracy of entrainment patterns. All the simulations that included any form of drag resulted in a fairly high density of particles entrained in a vortex. The entrainment and expulsion from eddies and vortices is an important characteristic of floating macroalgae. Accuracy in predicting entrainment and expulsion of macroalgae-rafts from eddies has been investigated previously by @putman2018 and @putman2020, who show that including windage greatly improves entrainment patterns when comparing virtual and real Sargassum rafts. However, the aforementioned studies investigated large rafts and not floating individuals. Inertia is based on the size/weight of the raft, while hydrodynamic and wind drag are based on current velocity and surface area. Therefore, the role of inertia may be negligible for a solitary individual and rather other forms of drag, such as in this study, should be considered. The results from this study suggest that the inclusion of hydrodynamic and/or wind drag is an important component when simulating virtual kelp particles will greatly increase the accuracy of entrainment and expulsion from sub-grid mesoscale oceanographic features, such as eddies and vortices.

The role of eddies and vortices in the entrainment of kelp is further supported by the accumulation simulations. These results cleary show that the inclusion of drag greatly increases the number of particles entrained in the eddy within the study domain. Eddies and vortices may act as accumulation zones for kelp where the formation of rafts may occur, or the kelp is entrained for long enough to eventually sink in area below the eddy. In a study by @woodborne1989 references a site offshore approximately 120km from Cape Columbine where dredged kelp was found. The area referenced by @woodborne1989 coincides with the area where virtual kelp are entrained by a vorticy. The use of a cylindrical shape coefficient instead of a sphere in the calculation of drag force also had no effect on the mean trajectory. Instead, the inclusion of a different shape coefficient increased to the density of particles entrained in an eddy field. This suggest that inclusion of different shape coefficients play a role in further improving estimates and hindcasts of macroalgal dispersal patterns.

@@

Most of the past research has focused on large macroalgae rafts which can vary greatly in size and shape, however no research exists attempting to clarify the drifting characteristics of solitary macroalgae. Understanding how to accurately model the distribution of solitary kelp can lead to be understanding of accumulation zones and sinks and ultimately the ecological links pertaining to those processes. The findings from this study indicate that solitary floating virtual *E. maxima* particles tend to follow the prevailing surface currents and the inclusion of wind causes the trajectories to differ greatly from that of a purely Lagrangian particle. Furthermore, the inclusion of both hydrodynamic and wind drag cause clustering of particles along the trajectory as well causing particles to become entrained in eddies. Different cross-sectional areas exposed to hydrodynamic and wind drag had no effect on overall trajectory suggesting that those differences are negligible when investigating dispersal patterns of drifting *E.maxima*. In addition, the use of a cylindrical shape coefficient has no effect on mean trajectory but also causes higher density of particles to become entrained in an eddy. Overall the inclusion of drag forces is an important aspect to consider when investigating the dispersal patterns of solitary macroalgae, while differences in cross-sectional areas are not. This study also provides an approach which can be adapted to model any floating solitary macroalgae, provided the surface area can be estimated accurately. Gaps in the research exist when considering what oceanographic processes play a role in solitary macroalgal dispersal as well as how these vary seasonally. The identification of biological and physical factors that play a role in accumulation zones is also needed, as well as better estimates for sinking rates and raft-times.

>>>>

Past research has inferred from direct and indirect techniques that drifting macroalgae tend to follow the prevailing surface currents [@thiel2005; @hobday2000; @fraser2011; @rothausler2011; @rothausler2011]. The results from this numerical experiment confirm that drifting macroalgae do follow the surface current in the study region when considering only hydrodynamic drag. In addition, not only are the trajectories similar but so are the end points at the end of simulation. However, when *ad-hoc* wind drag (windage) is considered in combination with hydrodynamic drag the end locations differ greatly when compared to purely Lagrangian particle trajectories. The inclusion of wind drag causes the kelp particles to flow further away from the coast and with a greater distance travelled from the release location. The difference in trajectory and distance covered can be attributed to the inclusion of wind drag. Wind drag adds to the effects of hydrodynamic drag when the wind direction and current direction are opposite to each other. However, when the wind direction is in the same or similar direction to that of the current, momentum energy from the wind is transferred to the object [@hackett2006; @putman2020]. This causes the kelp particles to travel further as well as exposed to different time varying flows. Past studies have shown this when including windage as a function of inertia [@putman2018;@brooks2019; @putman2020]. A study by @putman2020 used *ad-hoc* windage factors that was based on the surface current velocity to assess the approach for improving transport predictions of pelagic *Sargassum sp*. The results showed that including ad-hoc windage factors improved virtual trajectories compared to tracked *Sargassum sp* mats which was partly due to the inclusion of momemtum energy transfer from wind. This is also reflected in the analysis comparing distances travelled by all the particles in each simulation. The particles in the simulations that included wind drag traveled significantly greater distances compared to the passive or hydrodynamic only drag.

The addition of drag forces causes the kelp particles to cluster together along the mean trajectory compared to that of the passive particles which are more evenly dispersed. This is most likely due to the shared drag characteristics among the kelp particles which included the plants weight and surface area into the calculation. Other studies investigating flotsam trajectory characteristics have found similar results [@breivik2011; @olascoaga2020; @miron2020]. A study by @miron2020 conducted field experiments using a range of objects such as spheres, cubes, and cuboids to compare to investigate the effect of inertia on particles dynamics. The results from the study showed that objects tended to cluster according to the shape of the object which the authors attributed to shared characteristics of inertia. Although drag forces and not inertia were used in this study, the same conclusion applies; drag is therefore an important characteristic to consider when investigating or predicting solitary drifting macroalgae trajectory.

The ocean is comprised of time-varying flows and changes in an objects velocity that can cause the object to be exposed to different flow patterns over time. The lack of dissimilarity in trajectory between different cross-sectional area types (minimum, mean and maximum) suggests that reduction in velocity in relation to cross-sectional area is negligible. A similar result was found by @gouvello2020 who investigated the effects of swimming behavior on sea turtle hatchling dispersal in the Agulhas region. The results suggested that the hatchling trajectories are mostly influenced by the ocean currents in the first year of hatching due to  as juveniles do are not swim fast enough to alter thier overall trajectory. Therefore, the differences in cross-sectional area types may not be significant enough to have an effect on the relevant zonal and meridonal velocity vectors. Another possible reason is the resolution of the ocean model used. The resolution of the underlying ocean model is an important aspect of Lagrangian ocean modeling and must be able to resolve sub-grid scale ocean processes for finer scale applications. A study by @hart2018 assessed the inclusion of stochastic motion, wind and currents into forecasting for search and rescue using the same model ocean model used in this study. The authors found that the inclusion of Brownian motion greatly increased the accuracy in representing sub-grid scale processes and that the inclusion of wind, currents and stochastic motion greatly improved forecasting applications. Therefore, the authors argue that the lack of dissimilarity between simulations of different cross-sectional area and drag exposure scenarios is not as a result of the resolution of the model.

The same is also true when comparing different hydrodynamic and wind drag exposure scenarios. These scenarios reflect different buoyancy situations which result in different amounts of surface area exposed to hydrodynamic and/or wind drag. The similarities between the different combined drag scenarios suggest that higher magnitudes of wind exposure do not significantly alter trajectories. Instead, the results from this study suggest that the inclusion of drag forces in simulating macroalgal trajectory may result in improving accuracy of entrainment patterns. All the simulations that included any form of drag resulted in a  high density of particles entrained in a eddy. The entrainment and expulsion from eddies and vorticities are an important characteristic of floating macroalgae. Accuracy in predicting entrainment and expulsion of macroalgae-rafts from eddies has been investigated previously by @putman2018 and @putman2020, who show that including windage greatly improves entrainment patterns when comparing virtual and real Sargassum rafts. However, these studies investigated large rafts and not floating individuals. Inertia is based on the size/mass of the raft, while hydrodynamic and wind drag are based on current velocity and surface area. Therefore, the role of inertia may be negligible for a solitary individual and rather other forms of drag, such as in this study, should be considered. The results from this study suggest that the inclusion of hydrodynamic and/or wind drag is an important component when simulating virtual kelp particles, and will greatly increase the accuracy of entrainment and expulsion from sub-grid mesoscale oceanographic features, such as eddies and vortices.

The role of eddies and vortices in the entrainment of kelp is further supported by the accumulation simulations. These results cleary show that the inclusion of drag greatly increases the number of particles entrained in the eddy within the study domain. Eddies and vortices may act as accumulation zones for kelp where the formation of rafts may occur, or the kelp is entrained for long enough to eventually sink in area below the eddy. In a study by @woodborne1989 references a site offshore approximately 120km from Cape Columbine where dredged kelp was found. The area referenced by @woodborne1989 coincides with the area where virtual kelp are entrained by a vorticy. The use of a cylindrical shape coefficient instead of a sphere in the calculation of drag force also had no effect on the mean trajectory. Instead, the inclusion of a different shape coefficient increased to the density of particles entrained in an eddy field. Therefore, more accurately parameterizing shape will therefore allow for further improvements in models of macroalgal dispersal patterns.

@@

Most of the past research has focused on large macroalgae rafts which can vary greatly in size and shape; however, no research exists to clarify the drifting characteristics of solitary macroalgae. Understanding how to accurately model the distribution of solitary kelp can lead to be understanding of accumulation zones and sinks and ultimately the ecological links pertaining to those processes. The findings from this study indicate that solitary floating virtual *E. maxima* particles tend to follow the prevailing surface currents and the inclusion of wind causes the trajectories to differ greatly from that of a purely Lagrangian particle. Furthermore, the inclusion of both hydrodynamic and wind drag cause clustering of particles along the trajectory as well causing particles to become entrained in eddies. Different cross-sectional areas exposed to hydrodynamic and wind drag had no effect on overall trajectory suggesting that those differences are negligible when investigating dispersal patterns of drifting *E.maxima*. In addition, the use of a cylindrical shape coefficient has no effect on mean trajectory, but it does cause higher density of particles to become entrained in an eddy. Overall, the inclusion of drag forces is an important aspect to consider when investigating the dispersal patterns of solitary macroalgae, while differences in cross-sectional areas are not. This study also provides an approach which can be adapted to model any floating solitary macroalgae, provided the surface area can be estimated accurately. Gaps in the research exist when considering what oceanographic processes play a role in solitary macroalgal dispersal as well as how these vary seasonally. The identification of biological and physical factors that play a role in accumulation zones is also needed, as well as better estimates for sinking rates and raft-times.
